---
title: "Simple Random Sampling Equation Functions"
subtitle: "Elementary Survey Sampling, Chapter 4"
author: "Samuel Richards"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
    number_sections: no
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@scheaffer2011elementary

## Estimator for population mean

**(4.1)**,**(4.2)**
(@scheaffer2011elementary[p.83])

Estimator for population mean $\mu$:
$$
\hat{\mu} = \bar{y} =  \frac{1}{n}\sum_{i=1}^n y_i
$$

Estimator for the variance of $\bar{y}$
$$
\hat{V}(\bar{y}) = \left( 1-\frac{n}{N} \right) \frac{s^2}{n}
$$


## Estimator for population total

**(4.4)**,**(4.5)**
(@scheaffer2011elementary[p.87])

Estimator for population total $\hat{\tau}$:
$$
\hat{\tau} = N\bar{y} =  \frac{N}{n}\sum_{i=1}^n y_i
$$

Estimator for the variance of $\tau$
$$
\hat{V}(\hat{\tau}) = \hat{V}(N\bar{y}) = N^2 \left( 1-\frac{n}{N} \right) \frac{s^2}{n}
$$


### Function in R for population mean or total

```{r}
# n <- sample size
# y_i <- vector of observations
# N <- population size (if known)

pop_mean_or_total <- function(y_i, N, param){

n <- length(y_i)  #sample size
s2 <- var(y_i)    #sample variance

#----------Estimator for mu----------
  
  if (param == 'mu'){
    y_bar <- sum(y_i)/n   #population mean estimator
    
    Vy_bar <- (1-(n/N))*s2/n   #variance of the estimator

    B <- 2*sqrt(Vy_bar)   #Margin of error
    bounds <- y_bar + c(-B,B)   #Error bounds
  
#---return a list of parameter estimates---
  ybar_and_bounds <- list("Estimator for mean (y_bar)" = y_bar,
                "Variance of the estimator (V(y_bar))" = Vy_bar, 
                "Margin of Error (B)" = B,
                "Bounds on the error of Estimation" = bounds)
  return(ybar_and_bounds)
  }

#----------Estimator for tau----------
   
  else if (param == 'tau'){
    tau_bar <- N*sum(y_i)/n   #population total estimator
    
    Vtau_bar <- N^2*(1-(n/N))*s2/n   #variance of the estimator

    B <- 2*sqrt(Vtau_bar)   #Margin of error
    bounds <- tau_bar + c(-B,B)   #Error bounds

#---return a list of parameter estimates---
  taubar_and_bounds <- list("Estimator for total (tau_bar)" = tau_bar,
                "Variance of the estimator (V(tau_bar))" = Vtau_bar, 
                "Margin of Error (B)" = B,
                "Bounds on the error of Estimation" = bounds)
  return(taubar_and_bounds)
  }

}

# Test:

# y_i <- c(45,78,26,45,8,45,12,95,6,8,55)
# N <- 100
# pop_mean_or_total(y_i,N,param = "tau")
# pop_mean_or_total(y_i,N,param = "mu")
```



### A note on Bounds

The estimator for the bound on the error of estimation is $2\sqrt{V(\hat\lambda)}$ where $V(\hat\lambda)$ is the variance of the estimator ($\hat\mu$, $\hat\tau$, or $\hat{p}$).  
*(Notation deviates from the textbook)*  
$2\sqrt{}$ is an approximated version of a t-test that the text uses to simplify calculation.


## Sample size estimates for population mean and for population total

**(4.11)**,**(4.13)**
(@scheaffer2011elementary[p.88,90])

Sample size required to estimate $\mu$ or $\tau$ with a bound on the error of estimation $B$.
$$
n = \frac{N\sigma^2}{(N-1)D + \sigma^2} \\
\text{where} \\
D = \frac{B^2}{4} \text{ for } \mu \text{ and} \\
D = \frac{B^2}{4N^2} \text{ for } \tau
$$


### Fuction in R for sample size estimate

```{r}
# N <- population size
# s2 <- variance, often pulled from pilot study or prior experimentation
# B <- Desired margin of error

srs_samplesize <- function(N, s2, B, param){
  if(param == 'mu'){
    D <- B/4
  }
  else if(param == 'tau'){
    D <- B^2/(4*N^2)
  }
  
  n <- N*s2 / ((N-1)*D+s2)
  
  return(list("Approximate sample size:" = n))
}

#Test:

# N <- 1000
# s2 <- 78.5
# B <- 4
# srs_samplesize(N, s2, B, param = 'mu')
```


## Estimator for population proportion

**(4.14)**,**(4.15)**
(@scheaffer2011elementary[p.91])

Estimator for the population proportion $p$:
$$
\hat{p} = \bar{y} = \frac{\sum_{i=1}^n y_i}{n}
$$

Estimated variance of $\hat{p}$:
$$
\hat{V}(\hat{p}) = \left(1 - \frac{n}{N} \right) \frac{\hat{p} \hat{q}}{n-1} \\
\text{where} \\
\hat{q} = 1-\hat{p}
$$

### Function in R for population proportion

```{r}
# n <- sample size
# y_i <- vector of observations (as proportions)
# N <- population size (if known)

srs_pop_proportion <- function(y_i, N){

n <- length(y_i)  #sample size

#----------Estimator for p_hat----------
  
    p_hat <- sum(y_i)/n   #population proportion estimator
    
    q <- 1-p_hat    #calculation for proportion compliments
    
    Vp_hat <- (1-(n/N))*p_hat*q/(n-1)   #variance of the estimator

    B <- 2*sqrt(Vp_hat)   #Margin of error
    bounds <- p_hat + c(-B,B)   #Error bounds
  
#---return a list of parameter estimates---
  p_hat_and_bounds <- list("Estimator for population proportion (p_hat)" = p_hat,
                "Variance of the estimator (Vp_hat)" = Vp_hat, 
                "Margin of Error (B)" = B,
                "Bounds on the error of Estimation" = bounds)
  return(p_hat_and_bounds)

}

# Test:

# y_i <- c(45,78,26,45,8,45,12,95,6,8,55)/100
# N <- 100
# srs_pop_proportion(y_i,N)
```


## Sample size estimate for population proportion

**(4.18)**
(@scheaffer2011elementary[p.93])

Similar to Eq. **(4.11)** or **(4.13)**, $\sigma^2$ si replaced with $pq$ in the sample size formula to estimate $p$ with a bound on the error of estimation $B$:
$$
n = \frac{Npq}{(N-1)D + pq} \\
\text{where} \\
D = \frac{B^2}{4}
$$






