---
title: "Cluster Sampling Equation Functions"
subtitle: "Elementary Survey Sampling, Chapter 8"
author: "Samuel Richards"
date: "2023-02-27"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    number_sections: no
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@scheaffer2011elementary

## Notation
*Slide 12 for reference*  
(@scheaffer2011elementary[p.255])

$$
\begin{eqnarray}
N &-& \text{ number of clusters in the population} \\
n &-& \text{ number of clusters in a SRS} \\
m_i &-& \text{  number of elements in cluster } i \\
\bar{m} &=& \frac{1}{n} \sum_{i=1}^n m_i \text{ average cluster size for the sample} \\
M &=& \sum_{i=1}^N m_i \text{ number of elements in the population} \\
\bar{M} &=& \frac{M}{N} \text{ average cluster size for the population} \\
y_i &-& \text{ total of all observations in the } i^{th} \text{ cluster}\\
\end{eqnarray}
$$


## Population mean
*Slide 13 for reference*  
(@scheaffer2011elementary[p.255])

Estimator for the population mean $\mu$:
$$
\bar{y} = \frac{\sum_{i=1}^n y_i}{\sum_{i=1}^n m_i}
$$

Estimated variance of $\bar{y}$:
$$
\hat{V}(\bar{y}) = \left( \frac{N-n}{N n \bar{M}^2} \right) s_r^2 \\
\text{where} \\
s_r^2 = \frac{\sum_{i=1}^n (y_i - \bar{y} m_i)^2}{n-1}
$$
*Note:* If $\bar{M}$ is unknown, it can be approximated by $\bar{m}$.


### R function:

This function takes 
```{r}

```



## Population total
*Slide 21 for reference*  
(@scheaffer2011elementary[p.258])

Estimator for the population mean $\tau$:
$$
M\bar{y} = M \left( \frac{\sum_{i=1}^n y_i}{\sum_{i=1}^n m_i} \right)
$$

Estimated variance of $M\bar{y}$:
$$
\hat{V}(M\bar{y}) = M^2\hat{V}(\bar{y}) = N^2 \left( \frac{N-n}{Nn} \right) s_r^2 \\
\text{where} \\
s_r^2 = \frac{\sum_{i=1}^n (y_i - \bar{y} m_i)^2}{n-1}
$$


### R function:

This function takes 
```{r}

```



## Population total (M unknown)
*Slide 25 for reference*  
(@scheaffer2011elementary[p.259])

$$
N\bar{y_t} = \frac{N}{n} \sum_{i=1}^n y_i \text{ where } \bar{y_t} = \frac{\sum_{i=1}^n y_i}{n} 
$$

Estimated variance of $N\bar{y_t}$:
$$
\hat{V} (N\bar{y_t}) = N^2\hat{V}(\bar{y_t}) = N^2 \left( \frac{N-n}{Nn} \right) s_t^2 \\
\text{where} \\
s_t^2 = \frac{\sum_{i=1}^n (y_i - \bar{y_t})^2}{n-1}
$$


### R function:

This function takes 
```{r}

```



<!-- The following are not in the lecture notes but indeed are in the textbook -->

## Approximate sample size required to estimate population mean

**(8.12)**
(@scheaffer2011elementary[p.265])


Approximate sample size required to estimate $\mu$, with a bound $B$ on the error of estimation:
$$
n = \frac{N\sigma^2_r}{ND + \sigma^2_r} \\
\sigma^2_r \text{ is estimated by } s_r^2 \\
D = \frac{B^2\bar{M}^2}{4}
$$

## Approximate sample size required to estimate population total

**(8.13)**
(@scheaffer2011elementary[p.266])

Approximate sample size required to estimate $\tau$, using $M\bar{y}$, with a bound $B$ on the error of estimation:
$$
n = \frac{N\sigma^2_r}{ND + \sigma^2_r} \\
\sigma^2_r \text{ is estimated by } s_r^2 \\
D = \frac{B^2}{4N^2}
$$

(Note that the only difference between **(8.12)** and **(8.13)** is **D**.)


## Approximate sample size required to estimate population total (without M)

**(8.15)**
(@scheaffer2011elementary[p.267])

Approximate sample size required to estimate $\tau$, using $N\bar{y_t}$, with a bound $B$ on the error of estimation:
$$
n = \frac{N\sigma^2_t}{ND + \sigma^2_t} \\
\sigma^2_t \text{ is estimated by } s_t^2 \\
D = \frac{B^2}{4N^2}
$$


## Population proportion

**(8.16)**, **(8.17)**, **(8.18)**
(@scheaffer2011elementary[p.268])

Estimator for the population proportion $p$:
$$
\hat{p} = \frac{\sum_{i=1}^n a_i}{\sum_{i=1}^n m_i}
$$

Estimated variance of $\hat{p}$:
$$
\hat{V}(\hat{p}) = \left( \frac{N-n}{N n \bar{M}^2} \right) s_p^2 \\
\text{where} \\
s_p^2 = \frac{\sum_{i=1}^n (a_i - \hat{p} m_i)^2}{n-1}
$$

## Probabilities proportional to size

**(8.19)**, **(8.20)**, **(8.21)**, **(8.22)**
(@scheaffer2011elementary[p.275])

Estimator of the population mean $\mu$:
$$
\hat{u}_{pps} = \bar{\bar{y}} = \frac{1}{n}\sum_{i=1}^n \bar{y_i}
$$

Estimator for the variance of $\hat{u}_{pps}$:
$$
\hat{V}(\hat{u}_{pps}) = \frac{1}{n(n-1)} \sum_{i=1}^n (\bar{y_i} - \hat{u}_{pps})^2
$$

Estimator of the population total $\tau$:
$$
\hat{\tau}_{pps} = \frac{M}{n}\sum_{i=1}^n \bar{y_i}
$$

Estimator for the variance of $\hat{\tau}_{pps}$:
$$
\hat{V}(\hat{\tau}_{pps}) = \frac{M^2}{n(n-1)} \sum_{i=1}^n (\bar{y_i} - \hat{u}_{pps})^2
$$





